---
title: "Assignment 2: Physical Properties of Lakes"
author: "Tristen Townsend"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Hydrologic Data Analysis on the physical properties of lakes. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A02_LakePhysical.Rmd") prior to submission.

The completed exercise is due on 11 September 2019 at 9:00 am.

## Setup

1. Verify your working directory is set to the R project file, 
2. Load the tidyverse, lubridate, and cowplot packages
3. Import the NTL-LTER physical lake dataset and set the date column to the date format
4. Set your ggplot theme (can be theme_classic or something else)

```{r}

getwd()

library(tidyverse)
library(lubridate)  
library(cowplot)
library(RColorBrewer)

NTLdata <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")
NTLdata$sampledate <- as.Date(NTLdata$sampledate, "%m/%d/%y") 

theme_set(theme_classic())

```

## Creating and analyzing lake temperature profiles

### Single lake, multiple dates

5. Choose either Peter or Tuesday Lake. Create a new data frame that wrangles the full data frame so that it only includes that lake during two different years (one year from the early part of the dataset and one year from the late part of the dataset).

```{r}

#Choose Tuesday
Tuesdaydata <- filter(NTLdata, lakename == "Tuesday Lake")

#Filter for years 1985 and 2015
Tuesdaydata.1985.2015 <- filter(Tuesdaydata, year4 %in% c("1985", "2015"))

```

6. Create three graphs: (1) temperature profiles for the early year, (2) temperature profiles for the late year, and (3) a `plot_grid` of the two graphs together. Choose `geom_point` and color your points by date. 

Remember to edit your graphs so they follow good data visualization practices.

```{r}

#Temperature profile for 1985 
TuesdayTemp1985 <- 
  ggplot(filter(Tuesdaydata.1985.2015, year4 == 1985), aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point() +
  scale_y_continuous(breaks=seq(0, 17, 2.5)) +
  expand_limits(y=c(0,16)) +
  scale_y_reverse() +
  scale_x_continuous(position = "top") +
  expand_limits(x=c(0,25)) +
  scale_color_viridis_c(end = 0.8, option = "plasma") +  #end at 0.8 gets rid of yellow
  labs(x = expression("Temperature "(degree*C)), y = "Depth (m)") +
  theme(legend.position = "none")
print(TuesdayTemp1985)

#Temperature profile for 2015
TuesdayTemp2015 <- 
  ggplot(filter(Tuesdaydata.1985.2015, year4 == 2015), aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point() +
  scale_y_continuous(breaks=seq(0, 17, 2.5)) +
  scale_y_reverse() +
  expand_limits(y=c(0,16)) +
  scale_x_continuous(position = "top") +
  expand_limits(x=c(0,25)) +
  scale_color_viridis_c(end = 0.8, option = "plasma", name = "Ordinal Day") +  #end at 0.8 gets rid of yellow
  labs(x = expression("Temperature "(degree*C)), y = "") +
  theme(legend.position = "right", axis.text.y=element_blank())
print(TuesdayTemp2015)

Tuesday_cow <- 
  plot_grid(TuesdayTemp1985, TuesdayTemp2015, 
            ncol = 2, rel_widths = c(1, 1.4))
print(Tuesday_cow)

```

7. Interpret the stratification patterns in your graphs in light of seasonal trends. In addition, do you see differences between the two years?

> It appears that surface temperatures have increased.

### Multiple lakes, single date 

8. On July 25, 26, and 27 in 2016, all three lakes (Peter, Paul, and Tuesday) were sampled. Wrangle your data frame to include just these three dates.

```{r}

#Wrangling for just July 25, 26, 27
Julydata <- NTLdata %>% filter(sampledate == "2016-07-25" | sampledate == "2016-07-26" | sampledate == "2016-07-27")

```

9. Plot a profile line graph of temperature by depth, one line per lake. Each lake can be designated by a separate color.

```{r}

Julyplot <- ggplot(Julydata, aes(x=temperature_C, y=depth, color=lakename)) +
  geom_line() +
  scale_y_reverse() +
  scale_x_continuous(position = "top") +
  scale_color_brewer(palette="Paired", name="Lake Name") +
  labs(x = expression("Temperature "(degree*C)), y = "Depth (m)")
print(Julyplot)

```

10. What is the depth range of the epilimnion in each lake? The thermocline? The hypolimnion?

> The depth ranges of the epilimnion, thermocline, and hypolimnion are the surface to 1.3-2.5m, 2.5-6.25m, and 6.25 to the bottom, respectively.


## Trends in surface temperatures over time. 
11. Run the same analyses we ran in class to determine if surface lake temperatures for a given month have increased over time ("Long-term change in temperature" section of day 4 lesson in its entirety), this time for either Peter or Tuesday Lake. 

```{r}

#Add a month column to Tuesday dataframe
Tuesdaydata$Month <- month(Tuesdaydata$sampledate)

#Filter for surface depth
Tuesdaydata.surf <- Tuesdaydata %>% 
  filter(depth == 0.00)

#Create separate data frames for each month
Tuesday.may <- Tuesdaydata.surf %>% 
  filter(Month == 5)
  
Tuesday.june <- Tuesdaydata.surf %>% 
  filter(Month == 6)

Tuesday.july <- Tuesdaydata.surf %>% 
  filter(Month == 7)

Tuesday.aug <- Tuesdaydata.surf %>% 
  filter(Month == 8)

#Run linear regression
Tuesday.may.lm <- lm(temperature_C ~ year4, Tuesday.may) 
Tuesday.june.lm <- lm(temperature_C ~ year4, Tuesday.june) 
Tuesday.july.lm <- lm(temperature_C ~ year4, Tuesday.july) 
Tuesday.aug.lm <- lm(temperature_C ~ year4, Tuesday.aug) 

summary(Tuesday.may.lm) # p = 0.5558 DO NOT REJECT NULL
summary(Tuesday.june.lm) # p = 0.9921 DO NOT REJECT NULL
summary(Tuesday.july.lm) # p = 0.05691 DO NOT REJECT NULL
summary(Tuesday.aug.lm) # p = 0.1547 DO NOT REJECT NULL

```

12. How do your results compare to those we found in class for Paul Lake? Do similar trends exist for both lakes?

> For Paul Lake, we found that there were significant change in surface temperatures for July and August only for the time span reflected in the data. However, Tuesday Lake does not have any significant changes in surface temperatures.
