---
title: "Assignment 4: Water Quality in Rivers"
author: "Student Name"
output: html_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Hydrologic Data Analysis on water quality in rivers.

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single HTML file.
8. After Knitting, submit the completed exercise (HTML file) to the dropbox in Sakai. Add your last name into the file name (e.g., "A04_Chamberlin.html") prior to submission.

The completed exercise is due on 25 September 2019 at 9:00 am.

## Setup

1. Verify your working directory is set to the R project file, 
2. Load the tidyverse, dataRetrieval, cowplot, xts and dygraphs packages. 
3. Set your ggplot theme (can be theme_classic or something else)

```{r setup}

getwd()

library(tidyverse)
library(dataRetrieval)
library(cowplot)
library(xts)
library(dygraphs)

theme_set(theme_classic())

```

## Hypoxia

This assignment will look at another measure of water quality - oxygen concentration. Though not directly important for human health, oxygen in the water column is very important for aquatic life, and so is considered a measure of water quality. Hypoxia (low oxygen) has many different definitions. For this assignment, we will use 2 mg/L O~2~ as our cut-off.


4. Import the oxygen water quality data from New Hope Creek at Blands (using `readNWISqw()`, site code `02097314`, parameter code `00300`). Make a data frame called `O2.dat` that includes only the Date and O~2~ concentration values. Give your data frame understandable column names.

```{r read_data}

parameters <- parameterCdFile

O2.raw <- readNWISqw(siteNumbers = "02097314", #New Hope Creek at Blands 
                     parameterCd = c("00300"), 
                     startDate = "",
                     endDate = "")

O2.dat <- dat.raw %>%
  select(Date = sample_dt,
         Value = result_va) 

```

5. Create a ggplot of oxygen concentrations over time. Include a horizonal line at 2 mg/l to show the hypoxia cutoff. 

```{r}

O2_timeseries <- ggplot(O2.dat, aes(x = Date, y = Value)) +
  geom_point() +
  geom_hline(yintercept=2, color = "red", alpha = 0.8, lwd = 1, lty = 2)
print(O2_timeseries)

```

6. What do you notice about the frequency of hypoxia over time?

> Hypoxia only seemed to be an issue in the 1980s. After 1984 until the present there are no O2 sample values that indicate hypoxia is a problem in New Hope Creek at Blands. In general, the value of O2 has slightly decreased over the time series but has not resulted in hypoxic conditions (O2 < 2 mg/L).


## Nutrients

7. Often times hypoxia is associated with high nutrient concentrations, because abundant nutrients promote biomass growth which increases respiration and depletes oxygen concentrations in the water (remember how oxygen concentrations were very low in the hypolimnion from the Physical Properties of Lakes week). Create a new data frame, called `nutrients.dat` with total nitrogen (parameter code `00600`) and total phosphorus (parameter code `00665`) data from the USGS. Your data frame should have 3 columns, `Date`, `TotalNitrogen_mgl-N`, and `TotalPhosphorus_mgl-P`. 


```{r read_nutrients}

nutrients.dat.raw <- readNWISqw(siteNumbers = "02097314", #New Hope Creek at Blands 
                     parameterCd = c("00600", "00665"), 
                     startDate = "",
                     endDate = "")

nutrients.dat <- nutrients.dat.raw %>%
  select(Date = sample_dt,
         Parameter = parm_cd,
         Value = result_va) %>%
  mutate(id=1:n()) %>%
  spread(Parameter, Value) %>%
  select(-id) %>%
  rename(TotalNitrogen_mgl_N = '00600', TotalPhosphorus_mgl_P = '00665')

```

8. Create two ggplots stacked with cowplot that show nutrient concentrations over time.

```{r nutrient.plots)}

TN.plot <- ggplot(nutrients.dat) +
  geom_point(aes(x = Date, y = TotalNitrogen_mgl_N)) +
  #scale_color_manual(values = c("steelblue4", "darkorange4")) +
  ggtitle("Total Nitrogen, 1982-2018")
print(TN.plot)

TP.plot <- ggplot(nutrients.dat) +
  geom_point(aes(x = Date, y = TotalPhosphorus_mgl_P)) +
  #scale_color_manual(values = c("steelblue4", "darkorange4")) +
  ggtitle("Total Phosphorus, 1982-2018")
print(TP.plot)

plot_grid(TN.plot, TP.plot, label_size = 12)

```

9. What do these plots tell you about nutrient concentrations over time? How might this relate to your previous plot of hypoxia events?

> Nutrient concentrations were extremely high in the 1980s and sharpy decreased from around ~1985 to the present. This makes sense based on the previous hypoxia events that occured in the early to mid 1980s, since high nutrient concetrations were observed during that time as well.

## Discharge and Temperature

10. Turbulent flow in rivers mixes oxygen into the water column. As discharge decreases, water moves slower, and oxygen diffuses slower into the water from the atmosphere. Download and reformat the daily discharge data for New Hope Creek (function `readNWISdv()`, site `02097314`, parameter `00060`). 

```{r bring_discharge}

parameters <- parameterCdFile

discharge.raw <- readNWISdv(siteNumbers = "02097314", #New Hope Creek
                     parameterCd = c("00060"), 
                     startDate = "",
                     endDate = "")

discharge.dat <- discharge.raw %>%
  select(Agency = agency_cd,
         SiteCode = site_no,
         Date = Date, 
         Discharge.cfs = X_00060_00003, 
         ApprovalCode = X_00060_00003_cd) 

```

11. Create a dygraph of discharge and oxygen concentrations. You should have discharge on the y axis and oxygen concentration on the y2 axis. Set your y-axes limits so that they don't change as you zoom in and out. (hint: use `dyAxis(..., valueRange = c(0, 6500))` for discharge & `dyAxis(..., valueRange = c(0, 13))` for oxygen).
```{r}

O2 <- with(O2.dat, xts(x = Value, order.by = Date))
Discharge <- with(discharge.dat, xts(x = Discharge.cfs, order.by = Date))

DyDat <- cbind(O2, Discharge)

dygraph(DyDat) %>% 
  dySeries("O2", axis = "y2") %>% 
  dyAxis(name = "y", label = "Discharge (cfs)", valueRange = c(0, 6500)) %>%
  dyAxis(name = "y2", label = "Total O2 (cfu per 100 ml)", valueRange = c(0, 13)) %>%
  dyRangeSelector()

```

12. Do hypoxic events happen in the wet or the dry season? How many days after storm events (of at least 100 cfs) do they occur? 

> It appears the hypoxic events are happening in the dry season (August-October). They seem to occur approximately 75-85 days after the most recent storm event (of at least 100 cfs).

13. Temperature also impacts the solubility of oxygen in water. More oxygen can dissolve when temperatures are *lower*. How would you analyze the effects of temperature on dissolved oxygen concentrations? What data would you need? How would you separate the effects of temperature from the effects of flow?

> Though the hypoxic events occur during the summer months when temperatures are likely much higher, we see across the time series (when nutrient concentrations are much lower) that dissolved oxygen levels are much higher than the hypoxic value. 

14. What do you think is the biggest driver of hypoxia in New Hope Creek? What data do you use to support your conclusion?

> This leads us to think that the primary driver behind the hypoxic events are far more related to nutrient concentration and discharge rather than temperature.

## Reflection
15. What are 2-3 conclusions or summary points about river water quality you learned through your analysis?

>

16. What data, visualizations, and/or models supported your conclusions from 15?

> 

17. Did hands-on data analysis impact your learning about water quality relative to a theory-based lesson? If so, how?

>

18.	How did the real-world data compare with your expectations from theory?

>
